\Title

### 2013-02-01 synmut.tex
supplementary first draft

Intrapatient HIV evolution is dominated by selection on the protein level in the
arms race with the immune system. Synonymous mutations do not have an
immunity-related phenotype and are often assumed to be neutral. Here, we show
that synonymous changes in epitope-rich regions are often deleterious but still
reach high frequencies in the viral population.
We analyze longitudinal intra-patient data from the \shankaregion~part of the
envelope gene (\env) and observe that synonymous derived alleles rarely reach
fixation.
Simulations suggest that such synonymous mutations have a (Malthusian) selection
coefficient of the order of $-0.002$, and that they are brought up to high
frequency by hitchhiking on neighboring beneficial nonsynonymous alleles. We
detect a negative correlation between the fixation of an allele and its
involvement in RNA stem-loop structures.
The patterns of fixation of non-synonymous mutations suggest that antibody
escape mutations in \shankaregion~are only transiently beneficial, either since
the immune system is catching up or because of competition between equivalent
escapes.

@@

The central quantity we investigate is the probability of fixation of a
mutation, conditional on its population frequency.  A neutral mutation
segregating at frequency $\nu$ has a probability $P_\text{fix}(\nu) = \nu$ to
spread through the population and fix; in the rest of the cases, i.e.~with
probability $1-\nu$, it goes extinct. As illustrated in the inset of \FIG{aftsyn},
this is a simple consequence of the fact that
(a) exactly one of the $N$ individuals in the current population will be
the common ancestor of the entire future population at a particular locus and
(b) this ancestor has a probability $\nu$ of carrying the mutation (assuming
the neutral mutation is not preferentially associated with genomes of high or
low fitness).
Deleterious or beneficial mutations fix less or
more often than neutral ones, respectively. \FIG{aft}A\&B show 
the time course the frequencies of all synonymous and non-synonymous mutations
observed \env~, C2-V5, in patient p8~\citep{shankarappa_consistent_1999},
respectively. Despite many synonymous mutations reaching high frequency, very
few fix (panel~\ref{fig:aftsyn}); however, many nonsynonymous mutations fix
(panel~\ref{fig:aftnonsyn}).

@@

\begin{figure}
\begin{center}
\subfloat{\includegraphics[width=\linewidth]{simulations_graduallydel.pdf}
\label{fig:simfixpvar}}\\
\subfloat{\includegraphics[width=\linewidth]{simulations_syn.pdf}
\label{fig:simsfig}}
\caption{Distribution of selection coefficients on synonymous sites. Panel A)
The depression in $P_\text{fix}$ depends on the deleterious effect size 
of the synonymous alleles. This parameter also reduces synonymous
diversity, measured by the probability of a derived allele to be found at
intermediate frequencies $P_\text{interm}$ (first inset).
Panel B) To assess the parameter space that affects synonymous fixation and
diversity, we run many simulations with random parameters for deleterious effect
size, fraction of deleterious synonymous sites, average size of escape
mutations (color, blue ($10^{-2.5}$) to red ($10^{-1.5}$)), and rate of
introduction of new epitopes (marker size, from $10^{-3}$ to $10^{-2}$ per
generation). Mostly simulations with small deleterious effects and relatively
high deleterious fractions of synonymous mutations reproduce synonymous HIV-like
fixation probability and diversity. Parameters are chosen uniformly on a
logarithmic scale from the indicated windows. This simulation assumes 6
roughly equivalent escape mutations per epitope; see supplement for alternative
models. }
\label{fig:simheat}
\end{center}
\end{figure}

@@

Contrary to naive expectations, the adaptive escape mutations don't seem to be
unconditionally beneficial. Otherwise we should observe almost sure fixation of
non-synonymous mutations once they reach intermediate frequency. Instead, we
find that the fixation probability of non-synonymous mutations is roughly given
by its frequency. There are several possible explanations for this observation. 
Similar to synonymous mutations, the majority of non-synonymous mutations could
be weakly deleterious and the adaptive and deleterious parts conspire to yield
the neutral-like averaged fixation probability. We consider this possibility
inplausible since the amino-acid sequence outside the variable loops is much
more conserved that the synonymous positions, suggesting that the majority of
the non-synonymous mutations is much more deleterious.

Alternatively, the lack of fixation could be due to time dependent environment
through an immune system that is catching up, or competition between mutations
that mediate escape within the same epitope. We explored both of these
possibilities and find that both produce the desired effect. Furthermore, there
is evidence in support of both of these hypothesis. Serum from HIV infected
individuals tyically neutralizes the virus that dominated the population a few
(3-6) month ago \citep{richman_rapid_2003}. This suggests that escape mutations
cease to be beneficial after a few month and might revert if they come with a
fitness cost. Deep sequencing of regions of \env{} after antibody escape have
revealed multiple escape mutations in the same epitope
\citep{moore_limited_2009, bar_early_2012}. Presumably, each one of these
mutations is sufficient for escape but most combinations of them don't provide
any additional benefit to the virus. Hence only one of them will spread and the
others will be driven out of the population even if they transiently reach high
frequencies. The rapid emergence of multiple escape mutations in the same
epitope implies a large population size that explores all necessary point mutations rapidly. A
similar point has been made recently by Boltz {\it et al.} in the context of
preexisting drug resistance mutations~\citep{boltz_ultrasensitive_2012}.

@@

\section{Methods}
\subsection{Sequence data collection}
Longitudinal intrapatient viral RNA sequences were collected for published
studies~\citep{shankarappa_consistent_1999,
liu_selection_2006, bunnik_autologous_2008} and downloaded from the Los Alamos
National Laboratory (LANL) HIV sequence database~\citep{LANL2012}. The sequences from some patients
showed signs of HIV compartimentalization into subpopulations and were discarded; a total of 11
patients with approximately 6 time points each and 10 sequences per time point
were analyzed. The time interval or resolution between two consecutive sequences
was approximately 6 to 18 months.

>>>>

Intrapatient HIV evolution is dominated by selection on the protein level in the
arms race with the immune system. Synonymous mutations do not have an
immunity-related phenotype and are often assumed to be neutral. Here, we show
that synonymous changes in epitope-rich regions are often deleterious but still
reach high frequencies in the viral population.
We analyze longitudinal intra-patient data from the \shankaregion~part of the
envelope gene (\env) and observe that synonymous derived alleles rarely reach
fixation.
Simulations suggest that such synonymous mutations have a (Malthusian) selection
coefficient of the order of $-0.002$, and that they are brought up to high
frequency by hitchhiking on neighboring beneficial nonsynonymous alleles. We
detect a negative correlation between the fixation of an allele and its
involvement in RNA stem-loop structures.
The patterns of fixation of nonsynonymous mutations suggest that antibody
escape mutations in \shankaregion~are only transiently beneficial, either since
the immune system is catching up or because of competition between equivalent
escapes.

@@

The central quantity we investigate is the probability of fixation of a
mutation, conditional on its population frequency.  A neutral mutation
segregating at frequency $\nu$ has a probability $P_\text{fix}(\nu) = \nu$ to
spread through the population and fix; in the rest of the cases, i.e.~with
probability $1-\nu$, it goes extinct. As illustrated in the inset of \FIG{aftsyn},
this is a simple consequence of the fact that
(a) exactly one of the $N$ individuals in the current population will be
the common ancestor of the entire future population at a particular locus and
(b) this ancestor has a probability $\nu$ of carrying the mutation (assuming
the neutral mutation is not preferentially associated with genomes of high or
low fitness).
Deleterious or beneficial mutations fix less or
more often than neutral ones, respectively. \FIG{aft}A\&B show 
the time course the frequencies of all synonymous and nonsynonymous mutations
observed \env~, C2-V5, in patient p8~\citep{shankarappa_consistent_1999},
respectively. Despite many synonymous mutations reaching high frequency, very
few fix (panel~\ref{fig:aftsyn}); however, many nonsynonymous mutations fix
(panel~\ref{fig:aftnonsyn}).

@@

\begin{figure}
\begin{center}
\subfloat{\includegraphics[width=\linewidth]{simulations_graduallydel.pdf}
\label{fig:simfixpvar}}\\
\subfloat{\includegraphics[width=\linewidth]{simulations_syn.pdf}
\label{fig:simsfig}}
\caption{Distribution of selection coefficients on synonymous sites. Panel A)
The depression in $P_\text{fix}$ depends on the deleterious effect size 
of the synonymous alleles. This parameter also reduces synonymous
diversity, measured by the probability of a derived allele to be found at
intermediate frequencies $P_\text{interm}$ (first inset).
Panel B) To assess the parameter space that affects synonymous fixation and
diversity, we run many simulations with random parameters for deleterious effect
size, fraction of deleterious synonymous sites, average size of escape
mutations (color, blue ($10^{-2.5}$) to red ($10^{-1.5}$)), and rate of
introduction of new epitopes (marker size, from $10^{-3}$ to $10^{-2}$ per
generation). Mostly simulations with small deleterious effects and relatively
high deleterious fractions of synonymous mutations reproduce synonymous HIV-like
fixation probability and diversity. Parameters are chosen uniformly on a
logarithmic scale from the indicated windows. This simulation assumes 6
roughly equivalent escape mutations per epitope; see supplement for alternative
models.}
\label{fig:simheat}
\end{center}
\end{figure}

@@

Contrary to na\"ive expectations, the adaptive escape mutations don't seem to be
unconditionally beneficial. Otherwise we would observe almost sure fixation of
nonsynonymous mutations once they reach intermediate frequencies. Instead, we
find that the fixation probability of nonsynonymous mutations is roughly given
by its frequency. There are several possible explanations for this observation. 
Similar to synonymous mutations, the majority of nonsynonymous mutations could
be weakly deleterious and the adaptive and deleterious parts conspire to yield
the neutral-like averaged fixation probability. We consider this possibility
inplausible since the amino-acid sequence outside the variable loops is much
more conserved that the synonymous positions, suggesting that the majority of
the nonsynonymous mutations is much more deleterious.

Alternatively, the lack of fixation could be due to (a) time dependent environment
through an immune system that is catching up, or (b) competition between mutations
that mediate escape within the same epitope. We explored both of these
possibilities and find that both produce the desired effect. Furthermore, there
is evidence in support of both of these hypothesis. Serum from HIV infected
individuals tyically neutralizes the virus that dominated the population a few
(3-6) month ago \citep{richman_rapid_2003}. This suggests that escape mutations
cease to be beneficial after a few months and might revert if they come with a
fitness cost. Deep sequencing of regions of \env{} after antibody escape have
revealed multiple escape mutations in the same epitope
\citep{moore_limited_2009, bar_early_2012}. Presumably, each one of these
mutations is sufficient for escape but most combinations of them don't provide
any additional benefit to the virus. Hence only one mutation will spread and the
others will be driven out of the population although they transiently reach high
frequencies. The rapid emergence of multiple escape mutations in the same
epitope implies a large population size that explores all necessary point mutations rapidly. A
similar point has been made recently by Boltz {\it et al.} in the context of
preexisting drug resistance mutations~\citep{boltz_ultrasensitive_2012}.

@@

\section{Methods}
\subsection{Sequence data collection}

Longitudinal intrapatient viral RNA sequences were collected for published
studies~\citep{shankarappa_consistent_1999, liu_selection_2006,
bunnik_autologous_2008} and downloaded from the Los Alamos National Laboratory
(LANL) HIV sequence database~\citep{LANL2012}. The sequences from some patients
showed signs of HIV compartimentalization into subpopulations and were
discarded; a total of 11 patients with approximately 6 time points each and 10
sequences per time point were eventually analyzed. The time interval or
resolution between two consecutive sequences was approximately 6 to 18 months.
